Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-01-25 paper/zhang-cook-laa-langrene-menendez.tex
a coarse go-through of search_better -> SA etc, writing on polish, widefigure for trace plots

Below we introduce three line search algorithms in tour:
\texttt{search\_better}, \texttt{search\_better\_random}, and
\texttt{search\_geodesic}. The first two are simulated annealing
algorithms whilst \texttt{search\_geodesic} evaluates neighbourhood
basis to determine the search direction.

@@

\texttt{search\_better} is a random search device that samples a
candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the current
basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is a randomly generated matrix with the
same dimension as \(\mathbf{A}_{\text{cur}}\). \(\mathbf{A}_{l}\) is
then orthogonalised to ensure the orthonormal constraint is fulfilled.
When a basis is found with index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the search terminates and outputs the basis
for guided tour to construct an interpolation path. The next iteration
of search begins after adjusting \(\alpha\) by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\). The termination condition
is when the maximum number of iteration \(l_{\max}\) is reached. The
algorithm of \texttt{search\_better} is summarised in Algorithm
\ref{random-search}. A slightly different cooling scheme has been
proposed by \citet{posse1995projection} to avoid the search space being
reduced too fast. A halving parameter \(c\) is introduced and \(\alpha\)
is only adjusted if the last search takes more than \(c\) times to find
an accepted basis.

@@

Simulated annealing (\texttt{search\_better\_random})
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as \texttt{search\_better} but allows a
probabilistic acceptance of a basis with lower index value based on the
annealing \(T(l)\). Given an initial \(T_0\), the temperature at
iteration \(l\) is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). When a
candidate basis fails to have an index value larger than the current
basis, simulated annealing gives it a second chance to be accepted with
probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the algorithm to jump out of a local maximum and
enables a more holistic search of the whole parameter space. This
feature is particularly useful when local maxima are present. The
algorithm \ref{simulated-annealing} highlights how simulated annealing
differs from random search in the inner loop.

@@

\citet{cook1995grand} used a line search algorithm on the space of the
projection bases. In \texttt{search\_geodesic}, the search direction is
computed using the most prominent direction that deviating an tiny angle
of \(\delta\) from the current basis. The step size is chosen by
optimising the index value along the geodesic direction over an 90
degree angle from \(-\pi/4\) to \(\pi/4\) along the search direction
chosen. The optima \(\mathbf{A}_{**}\) is returned for the current
iteration if it meets the percentage improve condition or when
\(l_{\max}\) is reached. Algorithm \ref{search-geodesic} summarises the
inner loop in geodesic search.

@@

In the optimisation algorithms for projection pursuit, the three main
elements to record are 1) projection bases: \(\mathbf{A}\), 2) index
values: \(I\), and 3) State: \(S\), which labels the observation with
detailed stage in the optimisation. Possible values for
\texttt{search\_better} and \texttt{search\_better\_random} include
\texttt{random\_search}, \texttt{new\_basis}, and
\texttt{interpolation}. \texttt{search\_geodesic} has a wider variety
that includes \texttt{new\_basis}, \texttt{direction\_search},
\texttt{best\_direction\_search}, \texttt{best\_line\_search}, and
\texttt{interpolation}.

@@

Figure \ref{fig:toy-search} shows the searches of two different
optimisers: \texttt{search\_better} and \texttt{search\_better\_random}.
Both optimisers progress quickly at the first few iterations, take
longer to find better basis in the later iterations, and finish off when
hitting the maximum number of try, 499 and 999, respectively. Also, the
target basis found in each iteration of \texttt{search\_better} always
has an increased index value while this is not the case for
\texttt{search\_better\_random}. This explains why in this example,
\texttt{search\_better\_random} takes longer to find the final basis,
but in more complicated scenarios, the feature of probabilistic
acceptance allows a more holistic search of the basis space and is more
likely to find the global optimum.

@@

\caption{A comparison of search by two otimisers: \code{search\_better} and \code{search\_better\_random} on a six-variable dataset \code{boa6} with holes index on a 2D problem. Both optimisers finish when the maximum number of try is reached and attain similar index value in the end. \code{search\_better\_random} takes longer to get to the final basis because it allows for probabilistic acceptance of inferior basis, which is a feature useful for more complex problem.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

\caption[Trace plots of the bases on the interpolation path for two optimisers with the same configuration as the previous figure]{Trace plots of the bases on the interpolation path for two optimisers with the same configuration as the previous figure. The right path shows that to reach a lower target basis, the interpolation actually first passes some higher basis and this could potentially be information to be incorporated into the optimisers.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

@@

\caption[1D projection on the 5-variable dataset \code{boa5} with two optimisers]{1D projection on the 5-variable dataset \code{boa5} with two optimisers: pseudo derivative and simulated annealing. All the bases in pseudo derivative has been flipped positive so that the two paths finish close to each other and to the theoretical best.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

@@

For a particular index function, the best algorithm to optimise depends
on the character of the index and the data. If the index function is
smooth and has a single maximum, all of the three algorithms introduced
above can find the maximum. When multiple optima are present,
\texttt{search\_better} may get stuck at a local maximum. In the case
where the index function is non-smooth, \texttt{search\_geodesic} may
even fail to find the maximum. In this section, examples will be
presented to outline how the diagnostic plots can be used to compare the
performance of optimisers.

@@

A non-monotonic interpolation path has already been given in Figure
\ref{fig:toy-interp}: to get to the target basis from the current one,
the interpolator may have passed a basis with even higher index value.
This motivates the design of an interruption to check whether higher
index value can be reached during the interpolation. If such a basis can
be found, the optimiser will start the next iteration from that basis
instead of the target one. Figure \ref{fig:interruption} contrasts the
trace with and without the interruption for the simulated annealing
optimiser: \texttt{search\_better}. The first two interpolation are
identical while in the third interpolation, a better basis has been
found than the target one. Rather than starting the next iteration from
the ``target basis'' on Time 62, as shown in the left panel, the
interruption, on the right panel, starts the next iteration from the
``interpolation basis'' on Time 61. The interruption has also detected
better basis on the fourth interpolation. With the interruption, the
optimiser is able to find the final basis with a higher index value
(0.865 vs.~0.879).

@@

\caption[Trace plot of 2D projection on \code{boa6} data with holes index optimised by \code{search\_better}]{Trace plot of 2D projection on \code{boa6} data with holes index optimised by \code{search\_better}. The purpose of the plot is to compare the effect of implementing the interruption that avoid fully interpolating to the target basis if a better one can be found during the interpolation. With this implementation, the optimiser is able to find a better final basis.}\label{fig:interruption}
\end{widefigure}
\end{Schunk}

@@

\caption[Two-D projection on \code{boa6} data with holes index optimised by \code{search\_geodesic}]{Two-D projection on \code{boa6} data with holes index optimised by \code{search\_geodesic}. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish. }\label{fig:polish}
\end{figure}
\end{Schunk}

Once the final basis has been found by an algorithm, one may want to
push further to investigate whether there is an even better basis in the
close neighbourhood. This motivates the polish search where the final
basis is supplied as the start of a new guided tour to search for any
local breakthrough.

Similar to \texttt{search\_better} as a stochastic random search,
\texttt{search\_polish} has a different scheme of reducing the search
neighbourhood. In each search-interpolation iteration,
\texttt{search\_better} has a fixed neighbourhood parameter alpha and
this alpha is reduced by the cooling parameter only after an iteration
finishes. On the contrary, \texttt{search\_polish} allows alpha to be
reduced during each iteration to exploit the search in the
neighbourhood. Further, to avoid the case where alpha becomes too small
and the further search is meaningless, three more stopping criteria have
been added, on top of the original \texttt{max.tries} limit. These
include:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  the distance between the candidate basis and the current basis needs
  to be larger than 1e-3;
\item
  the percentage change of the index value need to be larger than 1e-5;
  and
\item
  the alpha parameter on itself needs to be larger than 0.01
\end{enumerate}

Figure \ref{fig:polish} presents the final projections found before and
after applying \texttt{search\_polish} on \texttt{search\_geodesic}.
Polish search improves the index value from 0.9618 to 0.9627 with
reduction of weights on the non-informative variables. In terms of the
projected data as in Figure \ref{fig:polish}, polish works to sharpen
the edges of each cluster.

@@

\[\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]\] With a more complex
index function, it is interesting to understand 1) if all the three
optimisers can reach the optimum, 2) when a local optimum is presented,
whether the simulated annealing algorithm can escape that local optimum
and find the global one.

@@

\caption[1D projection on 5\-variable dataset \code{boa5} with \code{norm\_kol} index]{1D projection on 5\-variable dataset \code{boa5} with \code{norm\_kol} index. The trace plot shows the failure of pseudo derivative on non-smooth index. The PCA plot also shows that simulated annealing is searching the area close to the theoretical best while pseudo derivative can't couple with the non-continuity in the index.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

@@

\caption[One-D projection of data \code{boa6} on \code{norm\_kol} index,  optimised by \code{search\_better} and \code{search\_better\_random}]{One-D projection of data \code{boa6} on \code{norm\_kol} index,  optimised by \code{search\_better} and \code{search\_better\_random}. Two different neighbourhood size of 0.5 and 0.7 is used. The finish point of each path is larger than the start point and optimums are separately labelled with * representing the global optimum and x the local optimum The color of the path is based on whether the optimiser finds the global or local optimum, after polishing.}\label{fig:kol-result}
\end{figure}
\end{Schunk}

The next experiment compares two optimisers: \texttt{search\_better} and
\texttt{search\_better\_random} for 20 simulation paths. Two search
neighbourhood sizes, \texttt{alpha}, of 0.5 and 0.7 are also compared to
understand the effect of neighbourhood size where a larger value
indicating a wider search. Figure \ref{fig:kol-result} shows 80 (20 * 4)
PCA projected interpolation paths faceted by optimiser and alpha.
Several results can be drawn through the comparison: 1) With
\texttt{search\_better} and \texttt{alpha\ =\ 0.5}, despite being the
simplest and fastest, the optimiser fails to optimise for three
instances where they finish neither near the local nor the global
optimum. 2) With the same optimiser but a larger neighbourhood size
\texttt{alpha\ =\ 0.7}, more seeds have found the global optimum. 3)
Comparing different optimisers \texttt{search\_better} and
\texttt{search\_better\_random} with the same search size of 0.5,
\texttt{search\_better\_random} finishes closer to the theoretical for
the ones that find the local optimum. This indicates the optimiser is
working harder to examine if there is other optimum in the space but a
small search neighbourhood size makes it hard to conduct a holistic
search of the whole space. 4) With \texttt{search\_better\_random} and a
larger neighbourhood size of 0.7, more seeds have found the global
optimum. Also some points are getting closer to the position of the
theoretical best, as comparing with \texttt{alpha\ =\ 0.5}, and this
would indicate fewer work for the subsequent polish.

>>>>

Below we introduce three line search algorithms in tour: simulated
annealing (SA), simulated annealing with jump out (SAJO), and pseudo
derivative (PD).

@@

Simulated annealing (SA) is a random search device that samples a
candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the current
basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is a randomly generated matrix with the
same dimension as \(\mathbf{A}_{\text{cur}}\). \(\mathbf{A}_{l}\) is
then orthogonalised to ensure the orthonormal constraint is fulfilled.
When a basis is found with index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the search terminates and outputs the basis
for guided tour to construct an interpolation path. The next iteration
of search begins after adjusting \(\alpha\) by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\). The termination condition
is when the maximum number of iteration \(l_{\max}\) is reached. The
algorithm of simulated annealing is summarised in Algorithm
\ref{random-search}. A slightly different cooling scheme has been
proposed by \citet{posse1995projection} to avoid the search space being
reduced too fast. A halving parameter \(c\) is introduced and \(\alpha\)
is only adjusted if the last search takes more than \(c\) times to find
an accepted basis.

@@

Simulated annealing with jump out (SAJO)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process but allows a probabilistic acceptance of a
inferior basis with lower index value based on the annealing \(T(l)\).
Given an initial \(T_0\), the temperature at iteration \(l\) is defined
as \(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to
have an index value larger than the current basis, simulated annealing
gives it a second chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the algorithm to jump out of a local maximum and
enables a more holistic search of the whole parameter space. This
feature is particularly useful when local maxima are present. The
algorithm \ref{simulated-annealing} highlights how SAJO differs from SA
in the inner loop.

@@

In pseudo derivative \citep{cook1995grand} method, the search direction
is computed using the most prominent direction that deviating an tiny
angle of \(\delta\) from the current basis. The step size is chosen by
optimising the index value along the geodesic direction over an 90
degree angle from \(-\pi/4\) to \(\pi/4\) along the search direction
chosen. The optima \(\mathbf{A}_{**}\) is returned for the current
iteration if it meets the percentage improve condition or when
\(l_{\max}\) is reached. Algorithm \ref{search-geodesic} summarises the
inner loop of the pseudo derivative search.

@@

Three main elements are recorded for the projection pursuit optimisers:
1) projection bases: \(\mathbf{A}\), 2) index values: \(I\), and 3)
State: \(S\), which labels the observation with detailed stage in the
optimisation. For optimiser SA and SAJO, possible values of the state
include \texttt{random\_search}, \texttt{new\_basis}, and
\texttt{interpolation}. pseudo derivative search has a wider variety of
state including \texttt{new\_basis}, \texttt{direction\_search},
\texttt{best\_direction\_search}, \texttt{best\_line\_search}, and
\texttt{interpolation}.

@@

Figure \ref{fig:toy-search} shows the searches of two different
optimisers: SA (left) and SAJO (right). Both optimisers progress quickly
at the first few iterations, take longer to find better basis in the
later iterations, and finish off when hitting the maximum number of try,
499 and 999, respectively. Also, the target basis found on the left
panel always has an increased index value in each iteration while this
is not the case for the plot on the right. This explains why in this
example, the optimiser SAJO takes longer to find the final basis, but in
more complicated scenarios, the feature of probabilistic acceptance
allows a more holistic search of the basis space and is more likely to
find the global optimum.

@@

\caption[A comparison of search by two otimisers]{A comparison of search by two otimisers: SA and SAJO on a six-variable dataset \code{boa6} with holes index on a 2D problem. Both optimisers finish when the maximum number of try is reached and attain similar index value in the end. Simulated annealing with jump out (SAJO) takes longer to get to the final basis because it allows for probabilistic acceptance of inferior basis, which is a feature useful for more complex problem.}\label{fig:toy-search}
\end{widefigure}
\end{Schunk}

@@

\caption[Trace plots of the bases on the interpolation path for two optimisers with the same configuration as the previous figure]{Trace plots of the bases on the interpolation path for two optimisers with the same configuration as the previous figure. The right path shows that to reach a lower target basis, the interpolation actually first passes some higher basis and this could potentially be information to be incorporated into the optimisers.}\label{fig:toy-interp}
\end{widefigure}
\end{Schunk}

@@

\caption[1D projection on the 5-variable dataset \code{boa5} with two optimisers]{1D projection on the 5-variable dataset \code{boa5} with two optimisers: Simulated Annealing (SA) and Pseudo Derivative (PD). All the bases in pseudo derivative has been flipped positive so that the two paths finish close to each other and to the theoretical best.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

@@

For a particular index function, the best algorithm to optimise depends
on the character of the index and the data. If the index function is
smooth and has a single maximum, all of the three algorithms introduced
above can find the maximum. When multiple optima are present, simulated
annealing (SA) may get stuck at a local maximum. In the case where the
index function is non-smooth, pseudo derivative search may even fail to
find the maximum. In this section, examples will be presented to outline
how the diagnostic plots can be used to compare the performance of
optimisers.

@@

A non-monotonic interpolation path has already been given in Figure
\ref{fig:toy-interp}: to get to the target basis from the current one,
the interpolator may have passed a basis with even higher index value.
This motivates the design of an interruption to check whether higher
index value can be reached during the interpolation. If such a basis can
be found, the optimiser will start the next iteration from that basis
instead of the target one. Figure \ref{fig:interruption} contrasts the
trace with and without the interruption for the simulated annealing
optimiser and it can be observed that while the first two interpolation
are identical, a better basis has been found than the target one in the
third interpolation. Rather than starting the next iteration from the
``target basis'' on Time 62, as shown in the left panel, the
interruption, on the right panel, starts the next iteration from the
``interpolation basis'' on Time 61. The interruption has also detected
better basis on the fourth interpolation. With the interruption, the
optimiser is able to find the final basis with a higher index value
(0.865 vs.~0.879).

@@

\caption[Trace plot of 2D projection on \code{boa6} data with holes index optimised by optimiser SA]{Trace plot of 2D projection on \code{boa6} data with holes index optimised by optimiser SA. The purpose of the plot is to compare the effect of implementing the interruption that avoid fully interpolating to the target basis if a better one can be found during the interpolation. With this implementation, the optimiser is able to find a better final basis.}\label{fig:interruption}
\end{widefigure}
\end{Schunk}

@@

\caption[Two-D projection on \code{boa6} data with holes index optimised by pseudo derivative]{Two-D projection on \code{boa6} data with holes index optimised by pseudo derivative. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish. }\label{fig:polish}
\end{figure}
\end{Schunk}

Once the final basis has been found by an optimiser, one may want to
push further in the close neighbourhood to find an even better basis. A
polish search takes the final basis as the start of a new guided tour to
search for local breakthrough.

The polish algorithm is similar to the simulated annealing one with
three main distinctions on 1) the number of candidate basis generated
each time in the inner loop; 2) the search neighbourhood,
\texttt{alpha}, and 3) the termination conditions. In simulated
annealing, only one candidate basis \(\mathbf{A}_l\) is generated every
time while the polish algorithm allows for a specified number of basis
to be evaluated for each repetition in the inner loop via an argument
\texttt{n\_sample}. The search neighbourhood in simulated annealing is
reduced in the outer loop, that is, the alpha for one iteration of
search remains the same, while in polish search, \texttt{alpha} is
reduced once a set of candidate bases fails to make an improvement. To
avoid the case the neighbourhood size becoming too small to generate
meaningful search, more termination criteria have been added, apart from
the original \texttt{max.tries} condition and these include:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  the distance between the basis found and the current basis needs to be
  larger than 1e-3;
\item
  the percentage change of the index value need to be larger than 1e-5;
  and
\item
  the alpha parameter on itself needs to be larger than 0.01
\end{enumerate}

Figure \ref{fig:polish} presents the final projections found before
(left) and after(right) applying the polish using the pseudo derivative
optimiser. Before polishing, the four clusters are clearly shown while
the edges of each cluster is relatively unshaved. The polish search
start with this basis and on the right panel, the edges of the clusters
are better trimmed.

@@

\[\max \left[F_{P}(u) - F_{y}(u)\right]\] With a more complex index
function, it is interesting to understand 1) if all the three optimisers
can reach the optimum, 2) when a local optimum is presented, whether the
simulated annealing algorithm can escape that local optimum and find the
global one.

@@

\caption[1D projection on 5\-variable dataset \code{boa5} with \code{norm\_kol} index]{1D projection on 5\-variable dataset \code{boa5} with \code{norm\_kol} index. The trace plot shows the failure of pseudo derivative on non-smooth index. The PCA plot also shows that simulated annealing is searching the area close to the theoretical best while pseudo derivative can't couple with the non-continuity in the index.}\label{fig:noisy-better-geo}
\end{widefigure}
\end{Schunk}

@@

\caption[1D projection of data \code{boa6} on \code{norm\_kol} index,  optimised by SA and SAJO]{1D projection of data \code{boa6} on \code{norm\_kol} index,  optimised by SA and SAJO. Two different neighbourhood size of 0.5 and 0.7 is used. The finish point of each path is larger than the start point and optimums are separately labelled with * representing the global optimum and x the local optimum The color of the path is based on whether the optimiser finds the global or local optimum, after polishing.}\label{fig:kol-result}
\end{figure}
\end{Schunk}

The next experiment compares the two simulated annealing optimisers, SA
and SAJO, for 20 simulation paths. Two search neighbourhood sizes,
\texttt{alpha}, of 0.5 and 0.7 are also compared to understand the
effect of neighbourhood size where a larger value indicating a wider
search. Figure \ref{fig:kol-result} shows 80 (20 * 4) PCA projected
interpolation paths faceted by optimiser and alpha. Several results can
be drawn through the comparison: 1) With optimiser SA and a 0.5
neighbourhood search size, despite being the simplest and fastest, the
optimiser fails to optimise for three instances where they finish
neither near the local nor the global optimum. 2) With the same
optimiser but a larger neighbourhood size of 0.7, more seeds have found
the global optimum. 3) Comparing between the optimiser SA and SAJO with
the same search size of 0.5, SAJO finishes closer to the theoretical for
the ones that find the local optimum. This indicates that it is working
harder to examine if there is other optimum in the space but a small
search neighbourhood size makes it hard to conduct a holistic search of
the whole space. 4) With optimiser SAJO and a larger neighbourhood size
of 0.7, more seeds have found the global optimum. Also some finish
points are getting closer to the position of the theoretical best, as
comparing with the 0.5 neighbourhood size, and this would indicate fewer
work for the subsequent polish.
